---
title: "Quick start of flowSpy"
author: "Yuting Dai"
package: flowSpy
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: github
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Quick start of flowSpy}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}
---

```{r use-libs, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE, 
                      warning = FALSE, message = TRUE)
```

This tutorial contains key steps of flowSpy workflow, including how to build an FSPY object, how to run clustering and dimension reduction, how to build an MST tree, how to run pseudotime and how to inference trajectory. It took about four minutes to run the whole analysis workflow with 12000 cells and 9 markers. 

If you want to see the detailed information of flowSpy workflow, please see the vignette **Tutorial of flowSpy**


``` {r quick-build, eval = FALSE}

# Loading packages
suppressMessages({
library(ggplot2)
library(LSD)
library(flowSpy)
library(LSD)
})

data("FSPYdata")

markers <- c("CD34", "CD43", "CD38", "CD90", "CD49f", "CD31", "CD45RA", "FLK1", "CD73")

# Create an FSPY object using fspy.meta.data and fspy.log.data
fspy <- createFSPY(raw.data = fspy.raw.data, 
                   markers = markers,
                   meta.data = fspy.meta.data,
                   log.transform = T)

# See information in object
fspy

# Run KNN
fspy <- runKNN(fspy, knn = 30)

# Run cluster based on SOM 
set.seed(1)
fspy <- runCluster(fspy, xdim = 6, ydim = 6)

# Run PCA
fspy <- runFastPCA(fspy)

# Run tSNE
fspy <- runTSNE(fspy)

# Run Diffusion Map
fspy <- runDiffusionMap(fspy)

# Run UMAP
fspy <- runUMAP(fspy)

# Build Trees using Minimum Spanning Tree (MST) 
fspy <- buildTree(fspy, cluster.type = "som", dim.type = "umap")

# Visualize Tree
plotTree(fspy, show.node.name = T, cex.size = 1.5)

# Define root cells
# The root id is from flowSpy tree, and is defined by users
fspy <- defRootCells(fspy, root.cells = c(34))

# Run pseudotime
fspy <- runPseudotime(fspy)

# Define leaf cells
# The root id is from flowSpy tree, and is defined by users
fspy <- defLeafCells(fspy, leaf.cells = c(4,9,29), pseudotime.cutoff = 0.5)

# Run walk from root cells to leaf cells
fspy <- runWalk(fspy)

# Show time point when the whole workflow is finished
Sys.time()

# Save object
save(fspy, file = "Path to you output directory")

```














